\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2013-03-25 arXiv_submission/smurf_paper.tex
Paper published. Updated and new version sent to arXiv.

% common acronyms etc.
\newcommand{\snr}{SNR}
\newcommand{\scuba}{SCUBA-2}
\newcommand{\rms}{RMS}

@@

Incoming light passes through a beam-splitter, and then bandpass
filters, providing simultaneous illumination and wavelength definition
of both the 450 and 850\,\micron\ focal planes. Each focal plan is
populated by four ``subarrays'' (labelled s4a--s4d at 450\,\micron, and
s8a--s8d at 850\,\micron), each consisting of 32 columns and 40 rows
of bolometers. The bolometers are thermal absorbers (with a response
time constant of $\tau \sim 1$\,ms) coupled to superconducting
transition-edge sensors (TESs) for thermometry. Temperature variations
in the TESs produce changing currents, and therefore varying magnetic
fields, which are detected and amplified using chains of
superconducting quantum interference devices (SQUIDs), before the
larger output currents are digitised. Each detector has its own SQUID
for the first-stage of the amplification, but the remaining stages
occur within a common chain of SQUIDs for each of the 32 columns. All
40 rows are read out in sequence at a row-visit frequency of about
12\,kHz. Such a high sample rate is unnecessary to produce maps, so
the data are re-sampled to approximately 200\,Hz before writing to
disk. This rate provides a sample every 3.0\,arcsec, or approximately
one third of the 450\,\micron\ diffraction-limited full-width at
half-maximum (FWHM) -- a typical rule-of-thumb for adequately sampling
a Gaussian point spread function (PSF) -- at the maximum scanning
speed of 600\,arcsec\,s$^{-1}$. There is an additional 41st row of
SQUID readouts that are not connected to TESs. These ``dark SQUIDs''
track non-thermal noise sources that are common to each column's
amplifier chain. The relationship between the output digitised
current, $I$, and the input power, $P$, or $dI/dP$ is established
using flatfield observations immediately prior to science
observations, in which the output signal is measured throughout a ramp
of the pixel heaters (which provide a known input power); see
Section~2.1 in \citet{dempsey2013} and Section~5.3 in
\citet{holland2013} for more details. Finally, the conversion to
astronomical flux units from power involves a correction for
atmospheric extinction \citep[primarily using the 183\,GHz (1.6\,mm)
JCMT Water Vapour Monitor to track line-of-sight opacity variations,
see Section~3 in][]{dempsey2013}, and the application of a flux
conversion factor (FCF) which is established from regular measurements
of calibrators such as Uranus \citep[Section~5 in][]{dempsey2013}.

@@

We note that the noise performance of \scuba\ has evolved over
time. An initial ``\scuba\ shared-risk observing'' period (S2SRO) took
place during February and March 2010, during which each of the 450 and
850\,\micron\ focal planes were populated with single subarrays, s4a
and s8d, respectively. In addition to having significantly fewer
available bolometers than the current fully-commissioned instrument
(and therefore a mapping speed reduced by a factor of $\sim$4),
instabilities in the fridge led to a large periodic signal with a
period of about 25\,s that was correlated amongst the detectors. After
upgrading and commissioning the instrument, a servo using newly added
thermometers in the focal planes has effectively mitigated this
problem \citep[Section~2.5 in][]{holland2013}. In addition, there were
improvements to the magnetic shielding (reducing magnetic field
pickup, as described in Section~\ref{sec:magpickup}), as well as more
effective removal of aliased noise sources (which has reduced both the
presence of line features and the mean white noise level).

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{magpickup}
\caption{Evidence for significant magnetic field pickup for
  observation 16 on 2010 February 28 at 450\,\micron\ (s4a subarray).
  The top panel shows two un-flatfielded (but mean-subtracted and step
  corrected) bolometer time-series from the same column, with a 200
  sample boxcar smooth (approximately 1\,s), illustrating that they
  are dominated by a similar signal with opposite signs. The second
  panel shows the dark squid signal for the column, also
  mean-subtracted and with the same boxcar smooth. The bottom panel
  shows the azimuthal and elevation offsets from the map centre (mean
  azimuth and elevation 171.9$^\circ$ and 68.0$^\circ$,
  respectively). Only the azimuthal signal is obviously correlated
  with the dark squids and bolometer signals, which suggests a
  magnetic field stationary with respect to the telescope dome as the
  source, since its direction with respect to the cryostat only
  changes with azimuthal motion.}
\label{fig:magpickup}
\end{figure}

@@

The basic method is as follows: (i) a covariance matrix is built up
for all pairs $(i,j)$ of the $N$ bolometer time-series,
$\langle\mathbf{b}_i(t),\mathbf{b}_j(t)\rangle$; and (ii) a singular
value decomposition identifies a new set of statistically uncorrelated
eigenvectors, $\mathbf{\xi}_i(t)$ (i.e., whose covariance matrix is
diagonal), such that each of the bolometer time-series is a linear
combination of the eigenvectors, or components i.e., $\mathbf{b}_i(t)
= \bar{\mathbf{\xi}} \mathbf{\lambda}_i^\mathrm{T}$, where each row of
the matrix $\bar{\mathbf{\xi}}$ is an eigenvector, and
$\mathbf{\lambda}_i^\mathrm{T}$ is a column vector containing the
corresponding eigenvalues. The $\mathbf{\xi}_i(t)$ are normalised by
$[\sum_t \mathbf{\xi}_i^2(t)]^{1/2}$, such that the amplitude of each
component in a given bolometer signal is stored in the eigenvalues. In
the earlier analyses mentioned, the low-frequency noise is assumed to
be encapsulated in those components with the largest
eigenvalues. Removing the projection of the time series along those
components then significantly reduces $1/f$ noise while retaining most
of the (higher-frequency) signal in point-like sources

A novel feature of our \scuba\ analysis is that we can perform PCA
with 450 and 850\,\micron\ data simultaneously, potentially helping us
to differentiate thermal and optical noise signals (e.g., from the
atmosphere) that might appear in both wavelengths, from other noise
sources that are restricted to single subarrays (such as readout
noise). In Fig.~\ref{fig:pca} we show the results for a combined
analysis of the s4a and s8b subarrays for the first $\sim$100\,s of
the same observation that was used in Figs.~\ref{fig:bolos_mix}b and
\ref{fig:pspec}b. The nine most significant components are shown
(ranked by the mean eigenvalues across all working bolometers in both
bands). The top and middle panels for each component show the
time-series and PSDs of the normalised eigenvectors. The bottom panels
are maps indicating the eigenvalues (amplitudes) of the component
across the focal plane (with the mean $\bar{\lambda}$ and and \rms,
$\lambda_\mathrm{rms}$, of the eigenvalues for all of the bolometers
calculated separately at each wavelength also shown).

The majority of the correlated signal at both wavelengths in
Fig.~\ref{fig:pca} is produced by Component~1, in which the
eigenvector time-series exhibits a roughly periodic signal that
resembles the scan pattern in Fig.~\ref{fig:bolos_mix}b. While this
correlated signal appears at both wavelengths, referring to the maps
of eigenvalues, they are considerably stronger at 850\,\micron\
(consistent with the visual appearance of the bolometer signals in
Fig.~\ref{fig:pca}). Also note that the PSD for this eigenvector
exhibits nearly a pure $1/f$ signature with minimal line features, and
no white-noise plateau (suggesting that these are high-\snr\
measurements of a purely low-frequency drift). Component~3 is weaker,
yet also shows evidence of the scan-synchronous noise, combined with
high-frequency line features that appear primarily in columns 26, 29
and 30 of the 450\,\micron\ subarray. Unlike Component~1, it is
generally more dominant at 450\,\micron.

\begin{figure*}
\centering
\includegraphics[width=0.85\linewidth]{pca}
\caption{The first nine components from a principal component
  analysis, ranked by the mean eigenvalues of combined 450 (s4a
  subarray) and 850\,\micron\ (s8b subarray) time-series bolometer
  data, for the same observation as that used in
  Figs.~\ref{fig:bolos_mix}b and \ref{fig:pspec}b. For each component
  the top plot shows the time-series of the eigenvectors normalised by
  their \rms, the middle plot its PSD, and the bottom coloured panels
  the eigenvalues for the bolometers at each wavelength (the
  amplitudes of the eigenvectors in the time-series). For reference,
  both the mean, $\bar{\lambda}$, and \rms, $\lambda_\mathrm{rms}$,
  eigenvalues for the bolometers in each subarray are also shown. Most
  of the correlated $1/f$ signal is encapsulated in Component~1, and
  appears to be dominated by scan-synchronous noise (compare with the
  scan pattern in Fig.~\ref{fig:bolos_mix}b). Some of this signal also
  appears in Component~3. Component~2 is another $1/f$ signal that is
  dominant in the 850\,\micron\ subarray. Most of the remaining
  components consist of weaker $1/f$ noise and line features
  predominantly above $\sim$30\,Hz. The eigenvalue maps show
  correlated patterns along columns suggesting pickup in their common
  amplifier chains.}
\label{fig:pca}
\end{figure*}

Why are there two components that seem to be dominated by
scan-synchronous noise (which is presumably a mixture of
magnetic field pickup and brightness variations in the atmosphere)?
First, the smoothly-varying gradients in the eigenvalue maps
(primarily at 850\,\micron) suggests that this noise has a different
response across the focal plane; in this case the PCA has identified
two orthogonal shapes that, when mixed in different quantities, can
reproduce most of the signal in each bolometer. It is also likely that
the atmospheric contribution is stronger at 450\,\micron\ (i.e.,
predominantly Component~3).

@@

\item if the block passes the above test, a least squares linear fit
is performed to the median-filtered bolometer data just before the
block, and this fit is extrapolated to predict the data value expected
at the centre of the block on the basis of the preceding data;

\item a least squares linear fit is performed to the median-filtered
bolometer data just after the block, and this fit is extrapolated to
predict the data value expected at the centre of the block on the
basis of the following data;

@@

\item Padding with artificial data: instead of padding with zeros,
each time-series is padded with artificial data that connects the two
ends of the data stream smoothly and includes Gaussian noise. No
apodisation is performed. The number of samples of padding at each end
is again equal to the ratio of the sampling frequency to the lowest
retained frequency.  This is illustrated in Fig.~\ref{fig:pad1}. See
\citet{stompor2002} for a thorough discussion of this procedure within
the context of CMB map-making. This method is used by default.

@@

\item As an alternative to the iterative map-making procedure,
cleaning by principal component analysis (PCA) is available as an
experimental pre-processing option (Section~\ref{sec:pca}). The most
significant components in the analysis are identified, and the
projection of each bolometer time-series along their eigenvectors are
removed. A single parameter specifies the threshold on the amplitude
of the eigenvalues to be removed, as a number of standard deviations
away from the mean value. Subarrays are cleaned independently, once,
for the full length of each continuous chunk of data. Given the
computational expense of this method, and initial tests which showed
little improvement over simple high-pass filtering, this method has
not yet been explored in detail with \scuba\ data. However, since
systematic effects seem to come and go, it is possible that PCA could
be useful for particular data sets.

@@

First, we describe the basic model for \scuba\ data. We express the
digitised signal observed by the $i$th bolometer as a function of
time,
%
\begin{equation}
\mathbf{b}_i(t) = f_i[\mathbf{e}_i(t) \mathbf{a}_i(t) + \mathbf{n}_i(t)],
\label{eq:model}
\end{equation}
%
where $\mathbf{a}(t)$ is the time-varying signal produced by scanning
the telescope across astronomical sources, $\mathbf{e}(t)$ is the
time-varying extinction, which is a function of the telescope
elevation and atmospheric conditions, and $\mathbf{n}_i(t)$ represents
sources of noise. The two terms in square brackets, as written, have
units of power delivered to the detectors (pW). The scale factor $f_i$
converts this effective power to the digitised units recorded on disk
(DAC) -- the flatfield multiplied by a digitisation constant (applied
once during pre-processing) -- which in this formulation is assumed to
be constant in time.

We then express the noise, $\mathbf{n}_i(t)$, as the sum of several
components,
%
\begin{equation}
  \mathbf{n}_i(t) = \mathbf{n}^\mathrm{w}_i(t) +
  g_i\mathbf{n}^\mathrm{c}(t) + \mathbf{n}^\mathrm{f}_i(t),
\label{eq:noise}
\end{equation}
%
where $\mathbf{n}^\mathrm{w}_i(t)$ is uncorrelated white noise,
$\mathbf{n}^\mathrm{c}(t)$ is a correlated or common-mode signal (with
an optional scale factor $g_i$ for each bolometer), and
$\mathbf{n}^\mathrm{f}_i(t)$ is (predominantly low-frequency) noise in
excess of the white noise level, that is either uncorrelated from
bolometer-to-bolometer, or at least does not have a simple correlation
relationship that would lead to it being included in
$\mathbf{n}^\mathrm{c}(t)$.

@@

A final option available as part of the \model{AST} model is to apply
constraints to the map to improve convergence, which presently include
setting user-specified or high-\snr\ regions to a value of zero for
all but the final iteration. See the discussion in
Section~\ref{sec:converge} and examples in
Sections~\ref{sec:examples}.

@@

To illustrate how large-scale ripples can form (and grow), the same
map solution is run for 100 iterations and shown in
Fig.~\ref{fig:pointmaps}b, now exhibiting a strong vertical
gradient. The degeneracy is easy to understand if the time-domain
behaviour of each model component is considered. The top panel of
Fig.~\ref{fig:degeneracy} shows the residual signals for a single
bolometer after 2 (black) and 100 (grey) iterations, which are nearly
identical, yet the change in the estimated
\model{COM}\footnote{\model{COM} has been multiplied by the
time-varying extinction correction to enable direct comparison with
\model{AST}.} (green) and \model{AST} (red) signals between 2 and 100
iterations are large However, it is also clear that the estimated
\model{COM} and \model{AST} signals are complimentary. In other words,
the large change in the \model{AST} signal is cancelled by freedom in
the \model{COM} signal to grow with opposite sign. For comparison, the
bottom panel of Fig.~\ref{fig:degeneracy} shows the telescope pointing
for this section of data, and the shapes of the \model{AST} and
\model{COM} signals match the elevation component, which is aligned
with the gradient in Fig.~\ref{fig:pointmaps}b. Generically, the
calculation of \model{COM} will remove any information on angular
scales that are larger than the array footprint (outline shown in
Fig.~\ref{fig:pointmaps}b for reference), meaning that the map
solution is unconstrained on such large scales.

@@

\scuba\ surveys designed to detect extremely faint point-sources
(e.g., high-redshift star-forming galaxies, and features in debris
disks) are ideally limited by the white-noise performance of the
instrument. The approach described here for maximising the \snr\ of
point-sources involves three major steps: (i) generating a map that
removes most low-frequency noise sources with approximately linear
response, without prior knowledge of the location of sources; (ii)
apply a Fourier-space ``whitening filter'' to suppress residual
large-scale noise; and (iii) detecting point sources using a ``matched
filter''. Note that variations on this general procedure have been
used extensively in the submm cosmology community using previous
instruments
\citep[e.g.,][]{scott2002,borys2003,laurent2005,coppin2006,scott2008,perera2008,devlin2009}.
In this section we reduce scans of the Lockman Hole taken during S2SRO
as a pilot project for the Cosmology Legacy Survey. It consists of
$\sim8.5$\,hours of data taken in 36 separate scans (average
15\,min. each) spread over February and March 2010. Each scan is a
360\,arcsec rotating PONG pattern, with a scan speed of
240\,arcsec\,s$^{-1}$, covering an area of about 50\,arcmin$^2$. The
full list of dates and observation numbers includes: 2010 February 18,
63, 64, 65, 70, 71, 72, 90, 91, 92, 97, 98, 99; 2010 February 20, 111,
112, 113, 118, 128, 129, 130; 2010 March 3, 61, 62, 64, 69, 70, 72,
73, 74, 75; 2010 March 9, 87, 88, 90, 91; and 2010 March 11, 59, 64, 65,
72. These observations represent about 80 per cent of the total data
taken as part of the project; the remaining observations were dropped
due to problems keeping the arrays properly tuned during S2SRO, and
were easily identified by their highly variable and erratic bolometer
time-series (which resulted in maps full of large streaks).

@@

For blind, high-redshift surveys, individual sources are expected to
be un-resolved by the \scuba\ 7.5--14.5\,arcsec FWHM beams. Under this
assumption, and also assuming that the map noise is white,
cross-correlation between the map and the known PSF, or matched
filtering, yields the maximum-likelihood flux density of supposed
point-sources centred over every location in the resulting map
\citep[an extremely well-known result throughout astronomy,
see][]{stetson1987}. Peak identification in such smoothed maps have
been used extensively in the submillimetre community, as both an
efficient source-detection and photometric measurement strategy.  For
the case at hand, we may use this cross-correlation technique since
the map has been whitened. However, we must first establish the
effective shape of point sources in this map due both to map-making
itself, and the whitening filter. We determine the effect of
map-making by adding a synthetic (and high-\snr) point source to the
real data, and measure its resulting shape in the map (solid and
dotted lines in Fig.~\ref{fig:lockman_psf}, respectively). Next, the
Fourier Transform of the map filtered PSF is divided by the square
root of the jackknife noise power spectrum to calculate the final
whitened PSF (dashed line in Fig.~\ref{fig:lockman_psf}). Maps
smoothed by this shape are shown for the whitened signal, and
jackknife maps in Figs.~\ref{fig:lockman_maps}c and f,
respectively. Note that these images are plotted in \snr\ units, where
the smooth noise maps have been calculated by propagating the original
noise maps output by SMURF both through the whitening and matched
filters (each of which are linear operations). In terms of the angular
power spectra, this complete process can be thought of as an optimal
band-pass filter that has both suppressed low-frequency noise, and
information on scales that are smaller (higher frequencies) than the
beam.

@@

Maps are produced using the first and second continuous halves of the
data in Fig.~\ref{fig:m17_jk}. This is not an ideal situation, since
the noise properties may evolve with time (e.g., due to changing sky
conditions), leading to a biased estimate of the parent noise
distribution in the complete map from the jackknife. Also, since the
zero-masking depends on the \snr\ of the map, it will be restricted to
regions approximately $\sqrt{2}$ shallower in these maps than for the
full data set. Finally, the cross-linking (position angles sampled) is
similar, though not identical in the two halves. Ideally one would
have many full maps, as in the case of the Lockman Hole data in the
previous section, from which alternating maps could be combined.

>>>>

% common acronyms etc.
\newcommand{\snr}{S/N}
\newcommand{\scuba}{SCUBA-2}
\newcommand{\rms}{rms}

@@

Incoming light passes through a beam-splitter, and then bandpass
filters, providing simultaneous illumination and wavelength definition
of both the 450 and 850\,\micron\ focal planes. Each focal plane is
populated by four ``subarrays'' (labelled s4a--s4d at 450\,\micron,
and s8a--s8d at 850\,\micron), each consisting of 32 columns and 40
rows of bolometers. The bolometers are thermal absorbers (with a
response time constant of $\tau \sim 1$\,ms) coupled to
superconducting transition-edge sensors (TESs) for
thermometry. Temperature variations in the TESs produce changing
currents, and therefore varying magnetic fields, which are detected
and amplified using chains of superconducting quantum interference
devices (SQUIDs), before the larger output currents are
digitised. Each detector has its own SQUID for the first-stage of the
amplification, but the remaining stages occur within a common chain of
SQUIDs for each of the 32 columns. All 40 rows are read out in
sequence at a row-visit frequency of about 12\,kHz. Such a high sample
rate is unnecessary to produce maps, so the data are re-sampled to
approximately 200\,Hz before writing to disk. This rate provides a
sample every 3.0\,arcsec, or approximately one third of the
450\,\micron\ diffraction-limited full-width at half-maximum (FWHM) --
a typical rule-of-thumb for adequately sampling a Gaussian point
spread function (PSF) -- at the maximum scanning speed of
600\,arcsec\,s$^{-1}$. There is an additional 41st row of SQUID
readouts that are not connected to TESs. These ``dark SQUIDs'' track
non-thermal noise sources that are common to each column's amplifier
chain. The relationship between the output digitised current, $I$, and
the input power, $P$, or $dI/dP$ is established using flatfield
observations immediately prior to science observations, in which the
output signal is measured throughout a ramp of the pixel heaters
(which provide a known input power); see Section~2.1 in
\citet{dempsey2013} and Section~5.3 in
\citet{holland2013} for more details. Finally, the conversion to
astronomical flux units from power involves a correction for
atmospheric extinction \citep[primarily using the 183\,GHz (1.6\,mm)
JCMT Water Vapour Monitor to track line-of-sight opacity variations,
see Section~3 in][]{dempsey2013}, and the application of a flux
conversion factor (FCF) which is established from regular measurements
of calibrators such as Uranus \citep[Section~5 in][]{dempsey2013}.

@@

We note that the noise performance of \scuba\ has evolved over
time. An initial ``\scuba\ shared-risk observing'' period (S2SRO) took
place during February and March 2010, during which each of the 450 and
850\,\micron\ focal planes were populated with single subarrays, s4a
and s8d, respectively. In addition to having significantly fewer
available bolometers than the current fully-commissioned instrument
(and therefore a mapping speed reduced by a factor of $\sim$4),
instabilities in the fridge led to a large oscillating signal with a
period of about 25\,s that was correlated amongst the detectors. After
upgrading and commissioning the instrument, a servo using newly added
thermometers in the focal planes has effectively mitigated this
problem \citep[Section~2.5 in][]{holland2013}. In addition, there were
improvements to the magnetic shielding (reducing magnetic field
pickup, as described in Section~\ref{sec:magpickup}), as well as more
effective removal of aliased noise sources (which has reduced both the
presence of line features and the mean white noise level).

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{magpickup}
\caption{Evidence for significant magnetic field pickup for
  observation 16 on 2010 February 28 at 450\,\micron\ (s4a subarray).
  The top panel shows two un-flatfielded (but mean-subtracted and step
  corrected) bolometer time-series from the same column, with a 200
  sample boxcar smoothing (approximately 1\,s), illustrating that they
  are dominated by a similar signal with opposite signs. The second
  panel shows the dark squid signal for the column, also
  mean-subtracted and with the same boxcar smoothing. The bottom panel
  shows the azimuthal and elevation offsets from the map centre (mean
  azimuth and elevation 171.9$^\circ$ and 68.0$^\circ$,
  respectively). Only the azimuthal signal is obviously correlated
  with the dark squids and bolometer signals, which suggests a
  magnetic field stationary with respect to the telescope dome as the
  source, since its direction with respect to the cryostat only
  changes with azimuthal motion.}
\label{fig:magpickup}
\end{figure}

@@

The basic method is as follows: (i) a covariance matrix is built up
for all pairs $(i,j)$ of the $N$ bolometer time-series,
$\langle\mathbf{b}_i(t),\mathbf{b}_j(t)\rangle$; and (ii) a singular
value decomposition identifies a new set of statistically uncorrelated
basis vectors, $\mathbf{\xi}_i(t)$ (i.e., whose covariance matrix is
diagonal), such that each of the bolometer time-series is a linear
combination of the basis vectors, or components i.e., $\mathbf{b}_i(t)
= \bar{\mathbf{\xi}} \mathbf{\lambda}_i^\mathrm{T}$, where each row of
the matrix $\bar{\mathbf{\xi}}$ is a basis vector, and
$\mathbf{\lambda}_i^\mathrm{T}$ is a column vector containing the
corresponding amplitudes. The $\mathbf{\xi}_i(t)$ are normalised by
$[\sum_t \mathbf{\xi}_i^2(t)]^{1/2}$, such that the amplitude of each
component is proportional to its \rms. In the earlier analyses
mentioned, the low-frequency noise is assumed to be encapsulated in
those components with the largest eigenvalues. Removing the projection
of the time series along those components then significantly reduces
$1/f$ noise while retaining most of the (higher-frequency) signal in
point-like sources

A novel feature of our \scuba\ analysis is that we can perform PCA
with 450 and 850\,\micron\ data simultaneously, potentially helping us
to differentiate thermal and optical noise signals (e.g., from the
atmosphere) that might appear in both wavelengths, from other noise
sources that are restricted to single subarrays (such as readout
noise). In Fig.~\ref{fig:pca} we show the results for a combined
analysis of the s4a and s8b subarrays for the first $\sim$100\,s of
the same observation that was used in Figs.~\ref{fig:bolos_mix}b and
\ref{fig:pspec}b. The nine most significant components are shown
(ranked by the mean amplitudes across all working bolometers in both
bands). The top and middle panels for each component show the
time-series and PSDs of the normalised basis vectors. The bottom
panels are maps indicating the significance (amplitudes) of the
component across the focal plane (with the mean $\bar{\lambda}$ and
and \rms, $\lambda_\mathrm{rms}$, of the amplitudes for all of the
bolometers calculated separately at each wavelength also shown).

The majority of the correlated signal at both wavelengths in
Fig.~\ref{fig:pca} is produced by Component~1, in which the basis
vector time-series exhibits a roughly periodic signal that resembles
the scan pattern in Fig.~\ref{fig:bolos_mix}b. While this correlated
signal appears at both wavelengths, referring to the maps of
amplitudes, they are considerably stronger at 850\,\micron\
(consistent with the visual appearance of the bolometer signals in
Fig.~\ref{fig:pca}). Also note that the PSD for this basis vector
exhibits nearly a pure $1/f$ signature with minimal line features, and
no white-noise plateau (suggesting that these are high-\snr\
measurements of a purely low-frequency drift). Component~3 is weaker,
yet also shows evidence of the scan-synchronous noise, combined with
high-frequency line features that appear primarily in columns 26, 29
and 30 of the 450\,\micron\ subarray. Unlike Component~1, it is
generally more dominant at 450\,\micron.

\begin{figure*}
\centering
\includegraphics[width=0.85\linewidth]{pca}
\caption{The first nine components from a principal component
  analysis, ranked by the mean amplitudes of combined 450 (s4a
  subarray) and 850\,\micron\ (s8b subarray) time-series bolometer
  data, for the same observation as that used in
  Figs.~\ref{fig:bolos_mix}b and \ref{fig:pspec}b. For each component
  the top plot shows the time-series of the basis vectors normalised
  by their \rms, the middle plot its PSD, and the bottom coloured
  panels the ampliudes for the bolometers at each wavelength (the
  contribution of the basis vectors to the time-series). For
  reference, both the mean, $\bar{\lambda}$, and \rms,
  $\lambda_\mathrm{rms}$, amplitudes for the bolometers in each
  subarray are also shown. Most of the correlated $1/f$ signal is
  encapsulated in Component~1, and appears to be dominated by
  scan-synchronous noise (compare with the scan pattern in
  Fig.~\ref{fig:bolos_mix}b). Some of this signal also appears in
  Component~3. Component~2 is another $1/f$ signal that is dominant in
  the 850\,\micron\ subarray. Most of the remaining components consist
  of weaker $1/f$ noise and line features predominantly above
  $\sim$30\,Hz. The amplitude maps show correlated patterns along
  columns suggesting pickup in their common amplifier chains.}
\label{fig:pca}
\end{figure*}

Why are there two components that seem to be dominated by
scan-synchronous noise (which is presumably a mixture of magnetic
field pickup and brightness variations in the atmosphere)?  First, the
smoothly-varying gradients in the amplitude maps (primarily at
850\,\micron) suggests that this noise has a different response across
the focal plane; in this case the PCA has identified two orthogonal
shapes that, when mixed in different quantities, can reproduce most of
the signal in each bolometer. It is also likely that the atmospheric
contribution is stronger at 450\,\micron\ (i.e., predominantly
Component~3).

@@

\item if the block passes the above test, a least squares linear fit
is performed to the median-filtered bolometer data just before the
block, and this fit is extrapolated to predict the data value expected
at the centre of the block;

\item a least squares linear fit is performed to the median-filtered
bolometer data just after the block, and this fit is extrapolated to
predict the data value expected at the centre of the block;

@@

\item Padding with artificial data: instead of padding with zeros,
each time-series is padded with artificial data which connects the two
ends of the data stream smoothly and includes Gaussian noise. No
apodisation is performed. The number of samples of padding at each end
is again equal to the ratio of the sampling frequency to the lowest
retained frequency.  This is illustrated in Fig.~\ref{fig:pad1}. See
\citet{stompor2002} for a thorough discussion of this procedure within
the context of CMB map-making. This method is used by default.

@@

\item As an alternative to the iterative map-making procedure,
cleaning by principal component analysis (PCA) is available as an
experimental pre-processing option (Section~\ref{sec:pca}). The most
significant components in the analysis are identified, and the
projection of each bolometer time-series along their basis vectors are
removed. A single parameter specifies the threshold on the amplitude
of the components to be removed, as a number of standard deviations
away from the mean value. Subarrays are cleaned independently, once,
for the full length of each continuous chunk of data. Given the
computational expense of this method, and initial tests which showed
little improvement over simple high-pass filtering, this method has
not yet been explored in detail with \scuba\ data. However, since
systematic effects seem to come and go, it is possible that PCA could
be useful for particular data sets.

@@

First, we describe the basic model for \scuba\ data. We express the
digitised signal observed by the $i$th bolometer as a function of
time,
%
\begin{equation}
\mathbf{b}_i(t) = f_i[\mathbf{e}_i(t) \mathbf{a}_i(t) + \mathbf{n}_i(t)],
\label{eq:model}
\end{equation}
%
where $\mathbf{a}(t)$ is the time-varying signal produced by scanning
the telescope across astronomical sources, $\mathbf{e}(t)$ is the
time-varying extinction, which is a function of the telescope
elevation and atmospheric conditions, and $\mathbf{n}_i(t)$ represents
sources of noise. The two terms in square brackets, as written, have
units of power delivered to the detectors (pW). The scale factor $f_i$
converts this effective power to the digitised units recorded on disk
-- the flatfield multiplied by a digitisation constant (applied once
during pre-processing) -- which in this formulation is assumed to be
constant in time.

We then express the noise, $\mathbf{n}_i(t)$, as the sum of several
components,
%
\begin{equation}
  \mathbf{n}_i(t) = \mathbf{n}^\mathrm{w}_i(t) +
  g_i\mathbf{n}^\mathrm{c}(t) + \mathbf{n}^\mathrm{f}_i(t),
\label{eq:noise}
\end{equation}
%
where $\mathbf{n}^\mathrm{w}_i(t)$ is uncorrelated white noise,
$\mathbf{n}^\mathrm{c}(t)$ is a correlated or common-mode signal (with
an optional scale factor $g_i$ for each bolometer), and
$\mathbf{n}^\mathrm{f}_i(t)$ is (predominantly low-frequency) noise in
excess of the white noise level, which is either uncorrelated from
bolometer-to-bolometer, or at least does not have a simple correlation
relationship that would lead to it being included in
$\mathbf{n}^\mathrm{c}(t)$.

@@

A final option available as part of the \model{AST} model is to apply
constraints to the map to improve convergence, which presently include
setting user-specified or low-\snr\ regions to a value of zero for
all but the final iteration. See the discussion in
Section~\ref{sec:converge} and examples in
Sections~\ref{sec:examples}.

@@

To illustrate how large-scale ripples can form (and grow), the same
map solution is run for 100 iterations and shown in
Fig.~\ref{fig:pointmaps}b, now exhibiting a strong vertical
gradient. The degeneracy is easy to understand if the time-domain
behaviour of each model component is considered. The top panel of
Fig.~\ref{fig:degeneracy} shows the residual signals for a single
bolometer after 2 (black) and 100 (grey) iterations, which are nearly
identical, yet the change in the estimated
\model{COM}\footnote{\model{COM} has been multiplied by the
time-varying extinction correction to enable direct comparison with
\model{AST}.} (green) and \model{AST} (red) signals between 2 and 100
iterations are large However, it is also clear that the estimated
\model{COM} and \model{AST} signals are complementary. In other words,
the large change in the \model{AST} signal is cancelled by freedom in
the \model{COM} signal to grow with opposite sign. For comparison, the
bottom panel of Fig.~\ref{fig:degeneracy} shows the telescope pointing
for this section of data, and the shapes of the \model{AST} and
\model{COM} signals match the elevation component, which is aligned
with the gradient in Fig.~\ref{fig:pointmaps}b. Generically, the
calculation of \model{COM} will remove any information on angular
scales that are larger than the array footprint (outline shown in
Fig.~\ref{fig:pointmaps}b for reference), meaning that the map
solution is unconstrained on such large scales.

@@

\scuba\ surveys designed to detect extremely faint point-sources
(e.g., high-redshift star-forming galaxies, and features in debris
disks) are ideally limited by the white-noise performance of the
instrument. The approach described here for maximising the \snr\ of
point-sources involves three major steps: (i) generating a map that
removes most large-scale noise sources with approximately linear
response, without prior knowledge of the location of sources; (ii)
apply a Fourier-space ``whitening filter'' to suppress residual
large-scale noise; and (iii) detecting point sources using a ``matched
filter''. Note that variations on this general procedure have been
used extensively in the submm cosmology community using previous
instruments
\citep[e.g.,][]{scott2002,borys2003,laurent2005,coppin2006,scott2008,perera2008,devlin2009}.
In this section we reduce scans of the Lockman Hole taken during S2SRO
as a pilot project for the SCUBA-2 Cosmology Legacy Survey. It
consists of $\sim8.5$\,hours of data taken in 36 separate scans
(average 15\,min. each) spread over February and March 2010. Each scan
is a 360\,arcsec rotating PONG pattern, with a scan speed of
240\,arcsec\,s$^{-1}$, covering an area of about 50\,arcmin$^2$. The
full list of dates and observation numbers includes: 2010 February 18,
63, 64, 65, 70, 71, 72, 90, 91, 92, 97, 98, 99; 2010 February 20, 111,
112, 113, 118, 128, 129, 130; 2010 March 3, 61, 62, 64, 69, 70, 72,
73, 74, 75; 2010 March 9, 87, 88, 90, 91; and 2010 March 11, 59, 64,
65, 72. These observations represent about 80 per cent of the total
data taken as part of the project; the remaining observations were
dropped due to problems in keeping the arrays properly tuned during
S2SRO, and were easily identified by their highly variable and erratic
bolometer time-series (which resulted in maps full of large streaks).

@@

For blind, high-redshift surveys, individual sources are expected to
be un-resolved by the \scuba\ 7.5--14.5\,arcsec FWHM beams. Under this
assumption, and also assuming that the map noise is white,
cross-correlation between the map and the known PSF, or matched
filtering, yields the maximum-likelihood flux density of supposed
point-sources centred over every location in the resulting map
\citep[an extremely well-known result throughout astronomy,
see][]{stetson1987}. Peak identification in such smoothed maps have
been used extensively in the submillimetre community, as both an
efficient source-detection and photometric measurement strategy.  For
the case at hand, we may use this cross-correlation technique since
the map has been whitened. However, we must first establish the
effective shape of point sources in this map due both to map-making
itself, and the whitening filter. We determine the effect of
map-making by adding a synthetic (and high-\snr) point source to the
real data, and measure its resulting shape in the map (solid and
dotted lines in Fig.~\ref{fig:lockman_psf}, respectively). Next, the
Fourier Transform of the map filtered PSF is divided by the square
root of the jackknife noise power spectrum to calculate the final
whitened PSF (dashed line in Fig.~\ref{fig:lockman_psf}).  Both the
whitened signal and jackknife maps are smoothed by this shape and
shown in Figs.~\ref{fig:lockman_maps}c and f, respectively. Note that
these images are plotted in \snr\ units, where the smoothed noise maps
have been calculated by propagating the original noise maps output by
SMURF through both the whitening and matched filters (each of which is
a linear operations). In terms of the angular power spectra, this
complete process can be thought of as an optimal band-pass filter that
has both suppressed low-frequency noise, and information on scales
that are smaller (higher frequencies) than the beam.

@@

Maps are produced using the first and second continuous halves of the
data in Fig.~\ref{fig:m17_jk}. This is not an ideal situation, since
the noise properties may evolve with time (e.g., due to changing sky
conditions), leading to a biased estimate of the parent noise
distribution in the complete map from the jackknife. Also, since the
zero-masking depends on the \snr\ of the map, it will be restricted to
regions approximately $\sqrt{2}$ times shallower in these maps than
for the full data set. Finally, the cross-linking (position angles
sampled) is similar, though not identical in the two halves. Ideally
one would have many full maps, as in the case of the Lockman Hole data
in the previous section, from which alternating maps could be
combined.
